FastSRM: A fast, memory efficient and identifiable implementation of the shared response model

### 2022-06-14 paper/main.tex
update stuff

What makes the analysis of fMRI data particularly difficult is that
the number of features (voxels) is usually much
larger than the number of available samples (time frames).
Yet, SRM has initially been designed to work within regions of interest using
only few subjects.
%
When using full brain data, computational costs become
high.
%
In addition, memory requirements are difficult to meet since the full dataset
needs to hold in memory.
%
Lastly, another issue of SRM is its non-identifiability (see the proof
in~\cite{richard2020modeling} Appendix D) which makes it challenging to
interpret the extracted shared components as they are defined up to a rotation.
%

In this work, we introduce FastSRM, an identifiable model that uses an optimal
spatial decomposition to speed up the computations with provably no loss of
performance.
%
FastSRM gains identifiability by imposing that the covariance of the shared
components is diagonal (see ~\cite{richard2020modeling} Appendix D for a proof).
%
The gains in speed and memory usage are based on the observation that a low-dimensional, yet distance-preserving
representation of the images yields the same result as the full data.
%
Such a representation can be interpreted as an initial spatial  decomposition of the data.
% that reduces the data in a such a way that the obtained
% shared components (and other parameters) are the same as if they were computed
% on full data.
%
We show that FastSRM brings several practical benefits: on real data, it produces more stable
estimates of the shared components than its non-identifiable counterpart, and 
 it is much faster and more memory
efficient than other implementations that do not make use of an optimal spatial decomposition.
% on synthetic as well as on real data,

@@

In~\cite{chen2015reduced} and~\cite{anderson2016enabling}, $\Sigma_s$ is only assumed to be definite positive.
%
As already highlighted in introduction, this causes the model to be
unidentifiable (see~\cite{richard2020modeling} Appendix D for a proof). Enforcing a diagonal $\Sigma_s$ ensures identifiability, provided that the diagonal values are different.
%
So we assume here that $\Sigma_s$ is diagonal (and refer the interested reader
to~\cite{chen2015reduced} and~\cite{anderson2016enabling} for the original formulation of Probabilistic SRM
without the diagonal constraint)
%

@@

\begin{figure}
  \centering
  \includegraphics[scale=0.34]{figures/srm/conceptual_figure2.pdf}
  \caption{\textbf{FastSRM algorithm} In step 1, data $\xb_i$ are projected onto a
    spatial decomposition $U_i$ that may depend on the subject $i$ (top).
    %
    In step 2 a SRM algorithm is applied on reduced data to compute the shared
    response.
    %
  }
  \label{fig:srm:conceptual}
\end{figure}

@@

\label{prop:optimaldetsrm}
\end{proposition}
The proof is available in appendix~\ref{proof:deterministic}.
In the case of probabilistic SRM we can obtain very similar results.
%
However the
algorithm applied on reduced data needs to be slightly modified.
%
We have the following result:
\begin{proposition}[Optimal spatial decomposition for probabilistic FastSRM]
  Let $(A_i)_i, \sigma_i, \Sigma_s$ be the solution obtained by probabilistic SRM on data
  $\xb_i$ and $(A_i')_i, \sigma_i', \Sigma_s'$ the solution obtained by ProbSRM on
  data $\zb_i = U_{\xb_i}^{\top} \xb_i$ but where updates:
  \begin{align}
    \sigma_i^2 \leftarrow \frac1{t} (\EE[\| \zb_i - A_i \EE[\sbb|\zb]\|^2] + \| \diag(\VV[\sbb | \zb]) \|^2)
  \end{align}
  are replaced by updates
  \begin{align}
    \sigma_i^2 \leftarrow \frac1{v} (\EE[\| \zb_i - A_i \EE[\sbb|\zb]\|^2] + \| \diag(\VV[\sbb | \zb]) \|^2)
  \end{align}
  %
  Then $A_i = U_{\xb_i}A_i'$, $\sigma_i =
  \sigma_i'$ and $\Sigma_s = \Sigma_s'$.
  %

@@

We consider SRM as a well principled formulation of the hyperalignment
method~\cite{haxby2011common}.
%
Many methods exist to perform hyperalignment: deep
hyperalignment~\cite{yousefnezhad2017deep} , robust
SRM~\cite{turek2018capturing} , multi-view canonical correlation
analysis~\cite{li2009joint}, ShICA~\cite{richard2021model},
MultiViewICA~\cite{richard2020modeling} optimal
transport~\cite{bazeille2019local} and many more.
%
In this work, we do not discuss the merit of FastSRM with respect to
other methods. We simply observe that SRM is widely used and aim to provide
a faster, more memory efficient and identifiable implementation.  In
our view, FastSRM should be used as a plugin replacement for SRM.

@@

FastSRM is an efficient implementation of SRM
that uses optimal spatial decompositions to speed up computations and reduce memory requirements
with provably no loss of performance. 
Note that the new memory requirements do not depend on the number of subjects.
This is really interesting as it would make it possible to apply FastSRM on
datasets with a very large number of subjects that do not hold in memory.

@@

FastSRM inherits from all the weaknesses of SRM including the fact that
mixing matrices are assumed to be orthogonal, which is debattable in practice.
%
It is an open question whether dimension reduction and learning of the shared
components could be done jointly and efficiently without assuming orthogonal
mixing matrices.
%
Our approach based on optimal spatial decomposition shows that SRM-like methods do not need the
full data to provide accurate estimates of shared components. We believe that such
insights may guide the design of future methods.

@@

\section{Detailed derivation of the Probabilistic SRM algorithm}
\label{details:probsrm}
Here we describe in detail the log-likeihood underlying the
Probabilistic SRM and derive the update rules.
%
Denoting $\VV[\sbb | \xb] = (\sum_i \frac1{\sigma_i^2} I +
\Sigma_s^{-1})^{-1}$ and $\EE[\sbb | \xb] = \VV[\sbb | \xb] \sum_i \frac1{\sigma_i^2}
A_i^{\top}\xb_i$, we have

@@

@@

  Then we consider the expected negative log-likelihood of probabilistic srm:
  \begin{align}
    \loss &= \sum_i \frac12 v\log(\sigma_i^2) + \frac12 \log(|\Sigma_s|) + \EE[\int_{\sbb} \sum_i \frac1{2 \sigma_i^{2}} \|\xb_i - A_i \sbb \|^2 \nonumber \\& \enspace \enspace \enspace \enspace+ \frac12 \dotp{ \sbb }{ \Sigma_s^{-1} \sbb }  d\sbb] \\
          &= \sum_i \frac12 v \log(\sigma_i^2) + \frac12 \log(|\Sigma_s|) + \EE[\int_{\sbb} \sum_i \frac1{2 \sigma_i^{2}}\|\zb_i - \tilde{A}_i \sbb \|^2 \nonumber \\& \enspace \enspace \enspace \enspace+ \frac12 \dotp{ \sbb }{ \Sigma_s^{-1} \sbb }  d\sbb]
  \end{align}
  where we use equality~\eqref{eq:equality:xy}.
  %
  %
  Optimizing the log-likelihood via expectation maximization yields the exact
  same updates as probabilistic SRM on data $\zb_i$
  except that updates
  \begin{align}
    \sigma_i^2 \leftarrow \frac1{t} (\EE[\| \zb_i - \tilde{A}_i \EE[\sbb|\zb]\|^2] + \| \diag(\VV[\sbb | \zb]) \|^2)
  \end{align}
  are replaced by updates
  \begin{align}
    \sigma_i^2 \leftarrow \frac1{v} (\EE[\| \zb_i - \tilde{A}_i \EE[\sbb|\zb]\|^2] + \| \diag(\VV[\sbb | \zb]) \|^2)
  \end{align}
  so that $\tilde{A}_i = A'_i$.
  %
  Therefore, the updates in both algorithms are linked  by $A_i = U_{\xb_i}A_i'$,
  $\sigma_i' =
  \sigma_i$ and $\Sigma_s'  = \Sigma_s$.
  %
  This concludes the proof.
  %
\end{proof}
\end{document}

>>>>

What makes the analysis of fMRI data particularly difficult is that
the number of features (voxels) is usually much
larger than the number of available samples (time frames).
Yet, SRM has initially been designed to work within regions of interest using
only few subjects.
%
When using full brain data, computational costs become
high.
%
In addition, memory requirements are difficult to meet since the full dataset
needs to hold in memory.
%
Lastly, another issue of SRM is its non-identifiability (see the proof
in~\cite{richard2020modeling} Appendix D also reported in Appendix~\ref{app:unidentifiability} for completness) which makes it challenging to
interpret the extracted shared components as they are defined up to a rotation.
%

In this work, we introduce FastSRM, an identifiable model that uses an optimal
spatial decomposition to speed up the computations with provably no loss of
performance.
%
FastSRM gains identifiability by imposing that the covariance of the shared
components is diagonal (see ~\cite{richard2020modeling} Appendix D or Appendix~\ref{app:unidentifiability} of this paper for a proof).
%
The gains in speed and memory usage are based on the observation that a low-dimensional, yet distance-preserving
representation of the images yields the same result as the full data.
%
Such a representation can be interpreted as an initial spatial  decomposition of the data.
% that reduces the data in a such a way that the obtained
% shared components (and other parameters) are the same as if they were computed
% on full data.
%
We show that FastSRM brings several practical benefits: on real data, it produces more stable
estimates of the shared components than its non-identifiable counterpart, and 
 it is much faster and more memory
efficient than other implementations that do not make use of an optimal spatial decomposition.
% on synthetic as well as on real data,

@@

In~\cite{chen2015reduced} and~\cite{anderson2016enabling}, $\Sigma_s$ is only assumed to be definite positive.
%
As already highlighted in introduction, this causes the model to be
unidentifiable (see~\cite{richard2020modeling} Appendix D or Appendix~\ref{app:unidentifiability} of this paper for a proof). Enforcing a diagonal $\Sigma_s$ ensures identifiability, provided that the diagonal values are different.
%
So we assume here that $\Sigma_s$ is diagonal (and refer the interested reader
to~\cite{chen2015reduced} and~\cite{anderson2016enabling} for the original formulation of Probabilistic SRM
without the diagonal constraint). The following paragraph gives update rules for $\lambda_i$ and $A_i$ and $\Sigma_s$. We highlight that the derivations used to obtain these updates are not new. They are obtained following the strategy introduced in~\cite{anderson2016enabling} and~\cite{chen2015reduced}.
%

@@

\begin{figure}
  \centering
  \includegraphics[scale=0.34]{figures/srm/conceptual_figure2_revised.pdf}
  \caption{\textbf{FastSRM algorithm} In step 1, data $\xb_i$ are projected onto a
    spatial decomposition $U_i$ that may depend on the subject $i$ (top).
    %
    In step 2 a SRM algorithm is applied on reduced data to compute the shared
    response.
    %
  }
  \label{fig:srm:conceptual}
\end{figure}

@@

\label{prop:optimaldetsrm}
\end{proposition}
The proof is available in appendix~\ref{proof:deterministic}.
In the case of probabilistic SRM we can obtain very similar results.
%
However the
algorithm applied on reduced data needs to be slightly modified.
%
We have the following result:
\begin{proposition}[Optimal spatial decomposition for probabilistic FastSRM]
  Let $(A_i)_i, \sigma_i, \Sigma_s$ be the solution obtained by probabilistic SRM on data
  $\xb_i \in \RR^v$ and $(A_i')_i, \sigma_i', \Sigma_s'$ the solution obtained by ProbSRM on
  data $\zb_i = U_{\xb_i}^{\top} \xb_i$, $\zb_i \in \RR^n$ but where updates:
  \begin{align}
    \sigma_i^2 \leftarrow \frac1{n} (\EE[\| \zb_i - A_i \EE[\sbb|\zb]\|^2] + \| \diag(\VV[\sbb | \zb]) \|^2)
  \end{align}
  are replaced by updates
  \begin{align}
    \sigma_i^2 \leftarrow \frac1{v} (\EE[\| \zb_i - A_i \EE[\sbb|\zb]\|^2] + \| \diag(\VV[\sbb | \zb]) \|^2)
  \end{align}
  %
  where $t$ is the number of samples and $v$ is the number of voxels.

  Then $A_i = U_{\xb_i}A_i'$, $\sigma_i =
  \sigma_i'$ and $\Sigma_s = \Sigma_s'$.
  %

@@

We consider SRM as a well principled formulation of the hyperalignment
method~\cite{haxby2011common}.
%
Many methods exist to perform hyperalignment: deep
hyperalignment~\cite{yousefnezhad2017deep} , robust
SRM~\cite{turek2018capturing} , multi-view canonical correlation
analysis~\cite{li2009joint}, ShICA~\cite{richard2021model},
MultiViewICA~\cite{richard2020modeling} optimal
transport~\cite{bazeille2019local} and many more.
%
SRM has been introduced after hyper-alignment in~\cite{chen2015reduced} and has been shown to outperform hyperalignment in many different settings. A recent work has come to a similar conclusion~\cite{bazeille2021empirical}. In this article, we do not discuss further the merit of FastSRM with respect to other methods. We simply observe that SRM is widely used and aim to provide
a faster, more memory efficient and identifiable implementation.  In
our view, FastSRM should be used as a plugin replacement for SRM.

@@

FastSRM is an efficient implementation of SRM
that uses optimal spatial decompositions to speed up computations and reduce memory requirements
with provably no loss of performance.

@@

FastSRM inherits from all the weaknesses of SRM including the fact that
mixing matrices are assumed to be orthogonal, which is debattable in practice.
Another desirable property of such decompositions is the spatial smoothness of the components. While it is often observed that SRM components are smooth, nothing imposes such a constraint in the model. We highlight that the optimal spatial decomposition is only optimal for the model we consider. Whether this representation is interpretable or not is irrelevant as it is only an intermediate step produced used during training.

%
It is an open question whether dimension reduction and learning of the shared
components could be done jointly and efficiently without assuming orthogonal
mixing matrices.
%
Our approach based on optimal spatial decomposition shows that SRM-like methods do not need the
full data to provide accurate estimates of shared components. We believe that such
insights may guide the design of future methods.

@@

\section{Detailed derivation of the Probabilistic SRM algorithm}
\label{details:probsrm}
Here we describe in detail the log-likeihood underlying the
Probabilistic SRM and derive the update rules.
We highlight that the derivations used to obtain these updates are not new. They are obtained following the strategy introduced in~\cite{anderson2016enabling} and~\cite{chen2015reduced}.

%
Denoting $\VV[\sbb | \xb] = (\sum_i \frac1{\sigma_i^2} I +
\Sigma_s^{-1})^{-1}$ and $\EE[\sbb | \xb] = \VV[\sbb | \xb] \sum_i \frac1{\sigma_i^2}
A_i^{\top}\xb_i$, we have

@@

\section{Identifiability results on Probabilistic SRM available in~\cite{richard2020modeling} appendix D}
\label{app:unidentifiability}
\begin{proposition}
Probabilistic SRM with positive definite $\Sigma_s$ is not identifiable
\end{proposition}
\begin{proof}
Let us assume the data $\xb_i \enspace i=1, \dots, m$ follow the Probabilistic SRM model with parameters $\Sigma_s, A_i, \sigma_i^2 \enspace i=1, \dots, m$ and $\Sigma_s$ definite positive.

Let us consider an orthogonal matrix $O \in \Ocal_k$.
We call $A'_i = A_i O$ and $\Sigma'_s = O^{\top} \Sigma_s O$.
$\Sigma'_s$ is trivially symmetric positive definite.

Then the data also follows the SRM model with different parameters $\Sigma'_s, A'_i, \sigma_i^2 \enspace i=1, \dots, m$.
\end{proof}

\begin{proposition}
We consider the Probabillistic SRM model where $\Sigma_s$ is a positive \emph{diagonal} matrix. We further assume that the values in $\Sigma$ are all distinct and ranked in ascending order.
Then, the Probabilistic SRM model is identifiable up to sign indeterminacies on the columns of
$\begin{bmatrix}
A^1 \\
\vdots \\
A^m
\end{bmatrix}
$.
\end{proposition}
\begin{proof}
The Probabilistic SRM can be written
\begin{align*}
    &\xb_i \sim \Ncal(0, A_i \Sigma_s {A_i}^{\top} + \sigma_i^2 I_v) \enspace \text{with}\enspace  {A_i}^{\top}A_i = I_p
\end{align*}
where $\Sigma_s$ is a positive diagonal matrix with distincts values ranked in ascending order.

Let us assume the data $\xb^i \enspace i=1, \dots, m$ follow the Probabilistic SRM model with parameters $\Sigma_s, A_i, {\sigma_i}^2 \enspace i=1, \dots, m$. Let us further assume that the data $\xb^i \enspace i=1, \dots, m$ follow the Probabilistic SRM model with an other set of parameters $\Sigma'_s, A'_i, {\sigma'_i}^2 \enspace i=1, \dots, m$.

Since the model is Gaussian, we look at the covariances.
We have for $i \neq j$
\begin{align*}
    \bbE[\xb_i\left(\xb_j\right)^{\top}] = A_i\Sigma_s {A_j}^{\top} = A'_i \Sigma'_s{A'_j}^{\top} \enspace,
\end{align*}
The singular value decomposition is unique up to sign flips and permutation. Since eigenvalues are positive and ranked the only indeterminacies left are on the eigenvectors. For each eigenvalue a sign flip can occur simultaneously on the corresponding left and right eigenvector.

Therefore we have $\Sigma'_s = \Sigma_s$, $A_i = A'_i D_{ij}$ and $A_j = A'_j D_{ij}$ where $D_{ij} \in \bbR^{k, k}$ is a diagonal matrix with values in $\{-1, 1\}$. This analysis holds for every $j \neq i$ and therefore $D_{ij} = D$ is the same for all subjects.

We also have for all $i$
\begin{align*}
    \bbE[\xb_i \left(\xb_i\right)^{\top}] = A_i \Sigma_s {A_i}^{\top} + \sigma_i^2 I_v =  A'_i \Sigma'_s {A'_i}^{\top}  + {\sigma}_i^2 I_v\\
\end{align*}
We therefore conclude ${\sigma}_i^2 = \sigma_i^2, i=1 \dots m$.

Note that if the diagonal subject specific noise covariance $\sigma_i^2 I_v$ is replaced by any positive definite matrix, the model still enjoys identifiability.
\end{proof}

@@

  Then we consider the expected negative log-likelihood of probabilistic srm:
  \begin{align}
    \loss &= \sum_i \frac12 v\log(\sigma_i^2) + \frac12 \log(|\Sigma_s|) + \EE[\int_{\sbb} \sum_i \frac1{2 \sigma_i^{2}} \|\xb_i - A_i \sbb \|^2 \nonumber \\& \enspace \enspace \enspace \enspace+ \frac12 \dotp{ \sbb }{ \Sigma_s^{-1} \sbb }  d\sbb] \\
          &= \sum_i \frac12 v \log(\sigma_i^2) + \frac12 \log(|\Sigma_s|) + \EE[\int_{\sbb} \sum_i \frac1{2 \sigma_i^{2}}\|\zb_i - \tilde{A}_i \sbb \|^2 \nonumber \\& \enspace \enspace \enspace \enspace+ \frac12 \dotp{ \sbb }{ \Sigma_s^{-1} \sbb }  d\sbb]
  \end{align}
  where we use equality~\eqref{eq:equality:xy}.
  %
  %
  Optimizing the log-likelihood via expectation maximization yields the exact
  same updates as probabilistic SRM on data $\zb_i$
  except that updates
  \begin{align}
    \sigma_i^2 \leftarrow \frac1{n} (\EE[\| \zb_i - \tilde{A}_i \EE[\sbb|\zb]\|^2] + \| \diag(\VV[\sbb | \zb]) \|^2)
  \end{align}
  are replaced by updates
  \begin{align}
    \sigma_i^2 \leftarrow \frac1{v} (\EE[\| \zb_i - \tilde{A}_i \EE[\sbb|\zb]\|^2] + \| \diag(\VV[\sbb | \zb]) \|^2)
  \end{align}
  so that $\tilde{A}_i = A'_i$.
  %
  Therefore, the updates in both algorithms are linked  by $A_i = U_{\xb_i}A_i'$,
  $\sigma_i' =
  \sigma_i$ and $\Sigma_s'  = \Sigma_s$.
  %
  This concludes the proof.
  %
\end{proof}
\end{document}
